---
title: "Project Report"
author: "Team Gamma"
date: 2021-10-14
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Central Question
What question did you set out to answer?



## Background
What is the relevant background on your question?
- Read up on and document background information on these data; you will have to provide background on the new data source in your report.


## Data 
What data did you find to help answer that question? 
  - Explore the dataset with EDA techniques, document your observations.
What is the quality of those data?
  - Comment on the quality of the data: How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?



You can include R code in the document as follows:

```{r cars}
summary(cars)
```

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

## Conclusions
What conclusions did you come to?

## Remaining Questions
What questions do you have remaining?

