---
title: "Project Report"
author: "Team Gamma"
date: 2021-10-14
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(gtrendsR)
library(tidyquant)
library(corrr)
library(ggcorrplot)
```

# Google Trends and the Pandemic

### Authors: Megan, Colin, Claire, Meagan, Akoua

## Introduction

In this project, our team set out to explore relationships between Google search trends and the state of the COVID-19 pandemic. Our central question was the following: "How does Google search activity correlate with the pandemic?"

## Background

### Google Trends

[Google Trends](https://trends.google.com/trends/?geo=US) provides Google search data, showing a measure of interest in a keyword and topic over time. Google trends anonymizes, categorizes, and groups together data, from which we can access a representative sample. It also normalizes search data by dividing each data point by total searches of location and time it represents and scaling numbers on a range of 0 to 100.

We pulled data from Google Trends to see how people react to different moments during the pandemic and explore the magnitude of those moments. When choosing specific search terms, we looked at the visual output on the Google Trends site to see if we can discover interesting patterns before bringing the data into R

To access Google Trends data without having to manually downloading .csv files for each query, we used the R package [`gtrendsR`](https://cran.r-project.org/web/packages/gtrendsR/gtrendsR.pdf) which scrapes the Google Trends site. We used `gtrends()` with different parameters for search keyword, time, and geographical location.

### New York Times COVID-19 Cases and Census Bureau

The New York Times has been collecting and publishing COVID case and death data since the first recorded case back in January 2020. The dataset is collected from local and state governments and health department in the Times' attempt to create a fuller picture of the pandemic. Data is reported by county and state and is publicly available via [GitHub](https://github.com/nytimes/covid-19-data).

We pair the above COVID data with population data from the [Census Bureau](https://www.census.gov/data.html), and follow the steps provided in c06. The population data can be found [here](https://data.census.gov/cedsci/table?q=United%20States&t=Population%20Total&g=0100000US%240500000&tid=ACSDT1Y2017.B01003&vintage=2017&layer=state&cid=DP05_0001E).

### Our World in Data Vaccination Rates

To look at vaccination rates, we used the Our World in Data vaccinations dataset, found on their [GitHub](https://github.com/owid/covid-19-data/tree/master/public/data/vaccinations). This data is collected by the **Our World In Data Team** based on official vaccination reports. According to their [site](https://ourworldindata.org/covid-vaccinations), "Our vaccination dataset uses the most recent official numbers from governments and health ministries worldwide. Population estimates for per-capita metrics are based on the United Nations World Population Prospects. Income groups are based on the World Bank classification."

## Exploratory Data Analysis

### NYT Covid Cases and Census Data

To import the COVID-19 case and death data, we will start by repeating the steps outlined in c06 by importing NYT data and Census data and creating one dataset.

```{r}
# Code taken from c06; credit to Zach del Rosario, professor at Olin College

df_pop <- read_csv(
  "./data/ACSDT5Y2018.B01003_data_with_overlays_2021-10-05T112340.csv", 
  skip = 1
)

## URL for the NYT covid-19 county-level data, raw
url_counties <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
filename_nyt <- "./data/nyt_counties.csv"

## Download the data locally
curl::curl_download(
        url_counties,
        destfile = filename_nyt
      )

## Loads the downloaded csv
df_covid <- read_csv(filename_nyt)

df_pop <- df_pop %>%
  mutate(fips = str_sub(id, -5))

df_covid <- df_covid %>%
  left_join(df_pop, "fips")

## Isolate specific columns
df_data <-
  df_covid %>%
  select(
    date,
    county,
    state,
    fips,
    cases,
    deaths,
    population = `Estimate!!Total`
  )

## Calculate case and death rates per 100k people
df_normalized <- df_data %>%
  mutate(
    total_cases_per100k = cases / population * 100000,
    total_deaths_per100k = deaths / population * 100000
  )

glimpse(df_normalized)

summary(df_normalized)
```
- Observations here!
-   Explore the dataset with EDA techniques, document your observations. What is the quality of those data?
-   Comment on the quality of the data: How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?


### Vaccination Data

```{r}
## URL for the OWID vaccination data
url_vax <- 
  "https://raw.githubusercontent.com/owid/covid-19-data/master/public/data/vaccinations/vaccinations.csv"
filename_vax <- "./data/owid_vax.csv"

## Download the data locally
curl::curl_download(
        url_vax,
        destfile = filename_vax
      )

## Loads the downloaded csv
df_vax <- read_csv(filename_vax) %>%
  filter(location == "United States") %>%
  drop_na(daily_vaccinations)

glimpse(df_vax)
summarize(df_vax)
```

- Observations here!
-   Explore the dataset with EDA techniques, document your observations. What is the quality of those data?
-   Comment on the quality of the data: How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?



### Google Trends Data

```{r}
# EDA goes here
```

- Observations here!
-   Explore the dataset with EDA techniques, document your observations. What is the quality of those data?
-   Comment on the quality of the data: How reputable is the source? How well-documented are their data-collection procedures? What potential errors might be in the data? What factors could contribute to missingness in the data?


What conclusions did you come to?

## Remaining Questions

What questions do you have remaining?
